Columnar VCF for Biobank scale datasets using Zarr

### 2024-03-29 paper.tex
Update abstract

% tentative
\title{Columnar VCF using Zarr}

@@

%% NOTE: this is much too long currently, but keeping for now so we
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
The largest dataset available today, the UK Biobank,
has a genotype matrix of approximately 1,000,000,000 rows and 500,000 columns,
along with metadata and quality control information of similar dimension.
Processing such matrices on a single computer is not feasible,
and computations must be distributed to complete in a reasonable timeframe.
Classical methods in genomics are inherently single-computer focused,
oriented around either streaming data through Unix pipelines or
storing the entire genotype matrix in memory.
The standard file format for genetic variation data (VCF: Variant Call Format)
is also fundamentally unsuited to distributed computation and is a severe
bottleneck at the Biobank scale.

\textbf{Results:}
key part of this approach is to use Zarr, a cloud-native
format for storing multi-dimensional numerical data, widely used in the
across the sciences to store petabyte-scale datasets, and with
multiple implementations.  We show that data stored in this format has
comparable compression levels to specialised methods to compress the
genotype matrix. We show that performing calculations on the genotype
matrix on a single computer is much more efficient at scale using sgkit than standard
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages.

\textbf{Conclusions:}
Such a standard, based on the language-agnostic Zarr format, would enable a
diverse ecosystem of tools to efficiently access subsets of the data without
copying, and substantially reduce research costs.
\end{abstract}

@@

\section{Background}

>>>>

% tentative
\title{Columnar VCF for Biobank scale datasets using Zarr}

@@

%% NOTE: this is much too long currently, but keeping for now so we
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
Variant Call Format (VCF) is the standard file format for interchanging
genetic variation data and associated quality control metrics.
It provides a well-defined data model and is central to a large ecosystem
of tools interoperating through this format.
The standard row-based encoding of the VCF data model (either as text
or packed binary) emphasises efficient retrieval of all data for a given
variant, but accessing subsets of the data within rows is inefficient.
Biobank-scale datasets currently available 
consist of hundreds of thousands of whole genomes with tens of terabytes
of compressed VCF per chromosome.
Row-based data storage is fundamentally unsuited to such datasets
and these large VCFs represent a major computational bottleneck.

\textbf{Results:}
We present the vcfzarr specification, a columnar encoding of the 
VCF data model using Zarr which makes retrieving subsets of the 
data much more efficient. Zarr is a cloud-native format for storing 
multi-dimensional numerical data, widely used in scientific computing.
We show how this format is competitive with specialised methods for 
storing genotype data in terms of compression ratios, and also performs
well on compressing the high-entropy quality-control fields that 
tend to dominate. We show that performing calculations on the genotype
matrix is much more efficient than standard
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages. We demonstrate 
the vcfzarr format (and the vcf2zarr conversion utility) 
on the Genomics England dataset of XX,XXX,XXX samples, showing
an overall reduction in storage of XX, and a reduction in processing 
time of common queries of up to XX fold.

\textbf{Conclusions:}
Large row-encoded VCF files are a major bottleneck for current research, and 
processing these files incurs a substantial economic cost.
The vcfzarr specification, building on widely-used, open-source technologies
has the potential to greatly reduce these costs,
and may enable a diverse ecosystem of next-generation tools for analysing 
genetic variation data.
\end{abstract}

@@

\section{Background}
[Four paragraphs. (1) What is VCF, and the good things it has done. 
(2) Explosion in dataset size. Summary of where we're at in terms of biobanks
etc. (3) We have a problem. Existing solutions like TileDB etc
are either proprietary, or tightly coupled to one specific software stack
(Hail, e.g.).
(4) We present the vcfzarr spec and the vcf2zarr conversion utility.]
